---
title: "Introduction to GIMP: Genomic Imprinting Methylation Patterns"
author: "Francesco Cecere and Abu Saadat"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to GIMP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

GIMP (Genomic Imprinting Methylation Patterns) is an R package for analyzing Imprinting Control Regions (ICRs) from DNA methylation array data. This package provides specialized tools for:

- Processing raw IDAT files from Illumina methylation arrays
- Extracting methylation data from Imprinting Control Regions
- Generating specialized visualizations for imprinting analysis
- Identifying differentially methylated positions in ICRs
- Interactive analysis through a Shiny web application

# Installation

To install GIMP from GitHub:

```{r install, eval=FALSE}
# Install devtools if not already installed
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

# Install GIMP
devtools::install_github("ngsFC/GIMP")
```

```{r load-package, eval=FALSE}
# Load the package
library(GIMP)
```

For IDAT file processing, additional Bioconductor packages are required:

```{r bioc-install, eval=FALSE}
# Install BiocManager if not already installed
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

# Install required annotation packages
BiocManager::install(c(
  "minfi",
  "IlluminaHumanMethylation450kanno.ilmn12.hg19",
  "IlluminaHumanMethylationEPICanno.ilm10b4.hg19",
  "IlluminaHumanMethylationEPICv2anno.20a1.hg38"
))
```

# Quick Start

## Using the Shiny Application

The easiest way to get started with GIMP is through the interactive Shiny application:

```{r shiny-app, eval=FALSE}
# Launch the interactive application
GIMP_app()

# For large files, increase upload limit
GIMP_app(max_upload_size_mb = 1000)
```

## Command Line Usage

### Working with Preprocessed Data

If you have preprocessed methylation beta values:

```{r preprocessed-example, eval=FALSE}
# Load your beta value matrix (samples as columns, probes as rows)
# beta_matrix <- readRDS("your_methylation_data.rds")

# Extract ICR-specific CpGs
ICRcpg <- make_cpgs(Bmatrix = beta_matrix, bedmeth = "v1")

# Create ICR-level summary
df_ICR <- make_ICRs(Bmatrix = beta_matrix, bedmeth = "v1")

# Define sample groups
sampleInfo <- c(rep("Control", 10), rep("Case", 10))

# Generate heatmap
ICRs_heatmap(df_ICR, sampleInfo = sampleInfo, plot_type = "beta")

# Identify differentially methylated positions
dmps <- iDMPs(data = ICRcpg, sampleInfo = sampleInfo)

# Plot specific ICR region
plot_line_ICR(significantDMPs = dmps$significantDMPs, 
              ICRcpg = ICRcpg, 
              ICR = "H19", 
              sampleInfo = sampleInfo)
```

### Processing Raw IDAT Files

For raw IDAT files stored in a ZIP archive:

```{r idat-example, eval=FALSE}
library(GIMP)

# Process IDAT files from ZIP archive
idat_data <- read_idat_zip(
  zip_file = "methylation_data.zip",
  array_type = "EPIC",
  normalize_method = "quantile"
)

# Extract the beta matrix and sample information
beta_matrix <- idat_data$beta_matrix
sample_info <- idat_data$sample_info

# Continue with standard GIMP workflow
ICRcpg <- make_cpgs(Bmatrix = beta_matrix, bedmeth = "v1")
df_ICR <- make_ICRs(Bmatrix = beta_matrix, bedmeth = "v1")

# Use sample groups from IDAT processing
sampleInfo <- sample_info

# Generate analysis
ICRs_heatmap(df_ICR, sampleInfo = sampleInfo)
```

# Array Type Selection

GIMP supports three main Illumina array types:

- **450k arrays**: Use `bedmeth = "450k"` (hg19 coordinates)
- **EPIC v1 arrays**: Use `bedmeth = "v1"` (hg19 coordinates)
- **EPIC v2 arrays**: Use `bedmeth = "v2"` (hg38 coordinates)

Always ensure your array type matches your data!

# Visualization Options

## Heatmap Types

GIMP provides three types of heatmaps:

1. **Beta values** (`plot_type = "beta"`): Raw methylation levels
2. **Delta values** (`plot_type = "delta"`): Difference from control mean
3. **Defect matrix** (`plot_type = "defect"`): Binary defect detection

```{r heatmap-types, eval=FALSE}
# Beta value heatmap
ICRs_heatmap(df_ICR, sampleInfo = sampleInfo, plot_type = "beta")

# Delta value heatmap  
ICRs_heatmap(df_ICR, sampleInfo = sampleInfo, plot_type = "delta")

# Defect matrix with custom threshold
ICRs_heatmap(df_ICR, sampleInfo = sampleInfo, plot_type = "defect", 
             sd_threshold = 2.5)
```

## Coverage Analysis

Visualize CpG coverage across ICRs:

```{r coverage, eval=FALSE}
# Plot CpG coverage
coverage_plots <- plot_cpgs_coverage(ICRcpg, bedmeth = "v1")

# View count plot
coverage_plots$plot_counts

# View percentage plot  
coverage_plots$plot_percentage
```

# Working with IDAT Files

## File Structure Requirements

Your ZIP file should contain:

```
methylation_data.zip
├── Sample1_ChipID_Position_Red.idat
├── Sample1_ChipID_Position_Grn.idat
├── Sample2_ChipID_Position_Red.idat
├── Sample2_ChipID_Position_Grn.idat
├── ...
└── samplesheet.csv
```

## Sample Sheet Format

The sample sheet must include these columns:

- `Sample_Name`: Unique sample identifier
- `Sentrix_ID`: Chip/slide ID
- `Sentrix_Position`: Position on chip (e.g., "R01C01")
- `Sample_Group`: Optional group assignment

```{r sample-sheet, eval=FALSE}
# Create a sample sheet template
template <- create_sample_sheet_template(
  sample_names = c("Control_01", "Control_02", "Case_01", "Case_02"),
  sentrix_ids = c("200123456", "200123456", "200123457", "200123457"),
  sentrix_positions = c("R01C01", "R02C01", "R01C01", "R02C01"),
  groups = c("Control", "Control", "Case", "Case")
)

# Save the template
write.csv(template, "samplesheet.csv", row.names = FALSE)
```

## Troubleshooting IDAT Issues

```{r idat-diagnostics, eval=FALSE}
# Preview ZIP contents before processing
preview_info <- preview_idat_zip("methylation_data.zip")

# Check minfi installation
check_minfi_functions()

# Validate ZIP structure
validation <- validate_idat_zip("methylation_data.zip")
```

# Advanced Analysis

## Custom Parameters

```{r advanced, eval=FALSE}
# Sensitive DMP detection
dmps_sensitive <- iDMPs(
  data = ICRcpg,
  sampleInfo = sampleInfo,
  pValueCutoff = 0.01  # More stringent
)

# Custom heatmap styling
ICRs_heatmap(
  df_ICR = df_ICR,
  sampleInfo = sampleInfo,
  plot_type = "defect",
  sd_threshold = 2.0,  # More sensitive defect detection
  order_by = "meth",   # Order by methylation
  annotation_col = list(Group = c("blue", "red"))
)
```

## Interactive Plotting

```{r interactive, eval=FALSE}
# Generate interactive line plot
plot_line_ICR(
  significantDMPs = dmps$significantDMPs,
  ICRcpg = ICRcpg,
  ICR = "MEST",
  sampleInfo = sampleInfo,
  interactive = TRUE  # Creates plotly interactive plot
)
```

# Session Information

```{r session-info, eval=FALSE}
sessionInfo()
```